---
title: 'Supplement to: Comment on Three Papers about Hardy-Weinberg Equilibrium Tests in Autopolyploids'
author: |
  | David Gerard
  | \small Department of Mathematics and Statistics, American University, Washington, DC, 20016, USA
output: 
  bookdown::pdf_document2:
    number_sections: true
    toc: false
bibliography: "geno_bib.bib"
urlcolor: "blue"
---

```{r setup, include=FALSE, message = FALSE}
set.seed(1)
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.align  = "center",
                      fig.height = 3, 
                      fig.width = 4)
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
options(digits = 3)
```

\begin{abstract}
This supplementary document contains additional simulations, coding examples, and other supporting material for ``Comment on Three Papers about Hardy-Weinberg Equilibrium Tests in Autopolyploids''.
\end{abstract}

\appendix
\newcommand{\bs}[1]{\boldsymbol{#1}}

\setcounter{table}{0}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{figure}{0}
\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{equation}{0}
\renewcommand{\theequation}{S\arabic{equation}}
\setcounter{section}{0}
\renewcommand{\thesection}{S\arabic{section}}


This document was written in R Markdown and may be explored interactively. All code chunks are executable in the order given. You can access the R Markdown file at <https://doi.org/10.5281/zenodo.7019205>.

The package `{hexocto}` contains the code from @wang2021tracing and @wang2022asymptotic, formatted in package form by me so that it is easier to compare. The original repos with the original code are <https://github.com/CCBBeijing/hexaploid> and <https://github.com/CCBBeijing/OctoploidDeer>. You can install this package using `{devtools}`:

``` r
# install.packages("devtools")
devtools::install_github("dcgerard/hexocto")
```

The package `{hwep}` contains the code from @gerard2022double. I use it for comparison purposes at times. You can install the development version via:

``` r
# install.packages("devtools")
devtools::install_github("dcgerard/hwep")
```

I will load these packages into R now:

```{r, message=FALSE}
library(hexocto)
library(hwep)
```
    
# Difference between random mating and equilibrium
\label{sec:diff}

Here, I numerically demonstrate the difference between random mating and equilibrium in autohexaploids.

For illustration, let's make an extreme example. Suppose the gamete frequencies for a hexaploid are
```{r}
p <- c(0, 0, 1, 0)
p
```

Then the genotype frequencies under random mating are

```{r}
q <- convolve(p, rev(p), type = "open")
round(q, digits = 3)
```

The allele frequency is
```{r}
r <- sum(0:6 * q) / 6
r
```

This results in equilibrium frequencies of the following when $\alpha = 0$, the lower bound,
```{r}
hwep::hwefreq(r = r, alpha = 0, ploidy = 6, niter = Inf)
```

I can verify this by iterating the recursive scheme from @wang2022asymptotic.
```{r}
qw <- q
for (i in 1:20) {
  qw <- hexocto::hex_onegen(yww = qw, alpha = 0)
}
qw
```

Equilibrium frequencies when $\alpha = 0.3$, the upper bound [@huang2019genotypic], are
```{r}
hwep::hwefreq(r = r, alpha = 0.3, ploidy = 6, niter = Inf)
```

I can also verify this by iterating the recursive scheme from @wang2022asymptotic.
```{r}
qw <- q
for (i in 1:20) {
  qw <- hexocto::hex_onegen(yww = qw, alpha = 0.3)
}
qw
```

# Incorrect equilibrium genotype frequencies from @wang2021tracing
\label{sec:wrong.equi}

I will begin at the same example genotype frequencies as @wang2021tracing.
```{r}
yww <- c(0.1, 0.1, 0.15, 0.1, 0.2, 0.1, 0.05, 0.1, 0.1)
```
Then I apply their recursive approach to obtain their equilibrium genotype frequencies
```{r}
hexocto::octo_recursive(yww = yww, niter = 20, alpha = 0)
```
These are different from the theoretical binomial proportions @haldane1930theoretical
```{r}
r <- sum(0:8 * yww) / 8
dbinom(x = 0:8, size = 8, prob = r)
```
My `{hwep}` package, on the other hand, correctly calculates these using my recursive formula
```{r}
qcurrent <- yww
for (i in seq_len(20)) {
  qcurrent <- hwep::freqnext(freq = qcurrent, alpha = c(0, 0))
}
qcurrent
```


# Coding errors for $\chi^2$ statistics
\label{sec:chi2.wrong}

```{r, echo = FALSE}
options(digits = 5)
```


@wang2022asymptotic use the following as an example for their tests for equilibrium and random mating on page 5 of their manuscript.

```{r}
nvec <- c(29, 21, 17, 10, 10, 10, 23)
nind <- sum(nvec)
```

Here, I will reproduce those tests, and demonstrate that they implemented their $\chi^2$ test statistics incorrectly.

Their recursive test gets a chi-squared value of 6.602, which I can get here.
```{r}
hex_chisq(yww = nvec / nind, 
          nind = nind,
          niter = 8,
          alpha = 0, 
          method = "incorrect")
```

This is the "incorrect" way because they forgot to account for the number of individuals in the chi-squared test. It should be 120 times 6.602.

```{r}
# generate their equilibrium frequencies
qhat <- hex_recursive(yww = nvec / nind, niter = 8, alpha = 0) 
# does not use nind
sum((qhat - (nvec / nind))^2 / (qhat)) 
# correct way
nind * sum((qhat - (nvec / nind))^2 / (qhat)) 
```

For the "gamete based test", they get 6.649, but this is not correct. They were just calculating the same test statistic as the 6.602 value, but ran it for a different number of iterations.

```{r}
# Estimate gamete frequencies
hout <- hex_em(yww = nvec / nind, niter = 30)
# Feed those into recursive algorithm
rvec <- hex_recursive(yww = hout$q, niter = 8, alpha = 0)
# Incorrect way
sum((nvec/nind - rvec)^2 / rvec)
```

Here is the value they were trying to get.
```{r}
# Incorrectly does not multiply by nind
sum((nvec/nind - hout$q)^2 / hout$q)
# Correctly multiplies by nind
nind * sum((nvec/nind - hout$q)^2 / hout$q)
```

The authors' two procedures would produce the same values if you ran them for long enough.
```{r}
# "recursive test" from Wang et al. (2021)
hex_chisq(yww = nvec / nind, 
          nind = nind,
          niter = 20,
          alpha = 0, 
          method = "incorrect")$chisq

# Implementation of "gamete-based" test from Wang et al. (2021)
hout <- hex_em(yww = nvec / nind, niter = 30)
rvec <- hex_recursive(yww = hout$q, niter = 20, alpha = 0)
sum((nvec/nind - rvec)^2 / rvec)
```

This is the exact same as just testing for binomial frequencies, but calculating the $\chi^2$ statistic incorrectly.

```{r}
rhat <- sum(nvec / nind * 0:6) / 6
qhat <- dbinom(x = 0:6, size = 6, prob = rhat)
sum((nvec/nind - qhat)^2 / qhat)
```

# Correct degrees of freedom

```{r, echo = FALSE}
options(digits = 3)
```

Here, I show that the method @wang2022asymptotic does not produce uniform $p$-values under the null of equilibrium. I also show that my correct version, including the correct degrees of freedom of 5, not 6, does produce uniform $p$-values under the null of equilibrium. I also find the correct degrees of freedom for the recursive test in @wang2021tracing to be 7, not 8.

## Hexaploids

I generate data under the null of equilibrium. I then fit the incorrect method @wang2022asymptotic, my corrected version, and the likelihood ratio test from @gerard2022double.
```{r}
qvec <- hwep::hwefreq(r = 0.5, alpha = 0.1, ploidy = 6)
nrep <- 1000
nsize <- 100000
pout_wang <- rep(NA_real_, length.out = nrep)
pout_correct <- rep(NA_real_, length.out = nrep)
pout_hwep <- rep(NA_real_, length.out = nrep)
for (i in seq_len(nrep)) {
  nvec <- c(rmultinom(n = 1, size = nsize, prob = qvec))
  pout_wang[[i]] <- hex_chisq(yww = nvec / sum(nvec),
                              nind = sum(nvec), 
                              alpha = 0.1,
                              method = "incorrect")$p
  pout_correct[[i]] <- hex_chisq(yww = nvec / sum(nvec),
                                 nind = sum(nvec),
                                 alpha = 0.1, 
                                 method = "correct")$p
  pout_hwep[[i]] <- hwep::hwelike(nvec = nvec, thresh = 0)$p_hwe
}
```

All of the $p$-values from @wang2022asymptotic are 1, so do not follow a uniform distribution.
```{r}
summary(pout_wang)
```

The QQ-plot of the correct $p$-values follow a uniform distribution.
```{r}
library(ggplot2)
qplot(sample = pout_correct, geom = "qq", distribution = qunif) +
  geom_abline()
```

The QQ-plot of the `{hwep}` $p$-values follow a uniform distribution.
```{r}
qplot(sample = pout_hwep, geom = "qq", distribution = qunif) +
  geom_abline()
```

## Octoploids

I generate data under the incorrect model of @wang2021tracing, calculating the $\chi^2$ statistic each iteration.
```{r}
yww1 <- c(0, 0, 0, 0, 1, 0, 0, 0, 0)
qvec <- octo_recursive(yww = yww1)
nrep <- 1000
nsize <- 100000
chstat_octo <- rep(NA_real_, length.out = nrep)
for (i in seq_len(nrep)) {
  nvec <- c(rmultinom(n = 1, size = nsize, prob = qvec))
  qemp <- nvec / sum(nvec)
  qnew <- octo_recursive(yww = qemp)
  chstat_octo[[i]] <- sum((qnew - qemp)^2 / qnew) * sum(nvec)
}
```

The degrees of freedom is not 8:
```{r}
pocto <- pchisq(q = chstat_octo, df = 8, lower.tail = FALSE)
qplot(sample = pocto, geom = "qq", distribution = qunif) +
  geom_abline()
```

The degrees of freedom *is* 7:
```{r}
pocto <- pchisq(q = chstat_octo, df = 7, lower.tail = FALSE)
qplot(sample = pocto, geom = "qq", distribution = qunif) +
  geom_abline()
```

# Simulation study to estimate $\alpha$
\label{sec:sim.alpha}

The model @wang2022asymptotic used to create an estimator for $\alpha$ is actually different from (1) and (2). Their model to estimate double reduction says that (i) parent genotypes frequencies satisfy $\tilde{\bs{q}} = \bs{p}*\bs{p}$ for some $\bs{p}$, and (ii) the current genotype frequencies are $\bs{q} = f(\tilde{\bs{q}}, \alpha)$. So this indicates random mating for parents, and one update of random mating for children given the double reduction rate.

I ran simulations with $\bs{p} = (1, 1, 1, 1)/4$ or $\bs{p} = (0.1, 0.2, 0.3, 0.4)$, $n \in \{100, 200, 400\}$, and $\alpha \in \{0, 1/7, 1/5, 3/11\}$. This mimics the simulation settings from @wang2022asymptotic. I ran each unique combination of parameter settings for 100 replications. Each replication, I generated data according to the assumed model from @wang2022asymptotic, then used their code to obtain estimates of $\bs{p}$ and $\alpha$. I always initialized the algorithm at $\alpha = 0$ and $\bs{p} = (1, 1, 1, 1) / 4$.

Below is my simulation code.

```{r, eval = FALSE}
## Parameter settings ----
pvec1 <- rep(1, 4) / 4
qvec1 <- convolve(pvec1, rev(pvec1), type = "open")
pvec2 <- c(0.1, 0.2, 0.3, 0.4)
qvec2 <- convolve(pvec2, rev(pvec2), type = "open")

niter <- 100
paramdf <- expand.grid(seed = seq_len(niter),
                       n = c(100, 200, 400),
                       alpha = c(0, 1/7, 1/5, 3/11),
                       truth = c("A", "B"))

## Estimates to fill in ----
paramdf$alphahat <- NA_real_
paramdf$p0hat <- NA_real_
paramdf$p1hat <- NA_real_
paramdf$p2hat <- NA_real_
paramdf$p3hat <- NA_real_

## Simulations ----
for (i in seq_len(nrow(paramdf))) {
  set.seed(paramdf$seed[[i]])
  ## offspring genotype frequencies
  if (paramdf$truth[[i]] == "A") {
    qoff <- hex_onegen(yww = qvec1, alpha = paramdf$alpha[[i]])  
  } else {
    qoff <- hex_onegen(yww = qvec2, alpha = paramdf$alpha[[i]])  
  }
  
  ## sample of offspring
  nvec <- c(rmultinom(n = 1, size = paramdf$n[[i]], prob = qoff))

  ## estimate parameters
  hout <- hex_estdr(NN = nvec, niter = 1000, tol = 0)
  paramdf$alphahat[[i]] <- hout$alpha
  paramdf$p0hat[[i]] <- hout$p[[1]]
  paramdf$p1hat[[i]] <- hout$p[[2]]
  paramdf$p2hat[[i]] <- hout$p[[3]]
  paramdf$p3hat[[i]] <- hout$p[[4]]
}
write.csv(x = paramdf, file = "./sims.csv", row.names = FALSE)
```

The estimates of $\alpha$ are very biased (Figure \@ref(fig:alpha-box)), and the estimates of $\bs{p}$ are somewhat biased (Figure \@ref(fig:p-box)). 

# Degrees of Freedom Calculations

Here, I list out the five instances of incorrect degrees of freedom calculations from @sun2021recursive, @wang2022asymptotic, and @wang2021tracing.

The degrees of freedom for the both the equilibrium and random mating
tests are incorrect in @sun2021recursive. They list the degrees
of freedom to be four in both tests. But there are already four free
parameters under the alternative (since
$q_0 + q_1 + q_2 + q_3 + q_4 = 1$). Since @sun2021recursive
assume the double reduction rate is known, under the null of
equilibrium there is one free parameter (the allele frequency), and so
the degrees of freedom for the test for equilibrium is $4-1=3$, not
4. Under the null of random mating, there are 2 free parameters (since
$p_0 + p_1 + p_2 = 1$), and so the degrees of freedom for the test of
random mating is $4 - 2 = 2$, not 4.

The degrees of freedom for the random mating test is incorrect in
@wang2022asymptotic. On page 4 of @wang2022asymptotic,
the authors say about their test for random mating that "this test
statistic follows the chi-square distribution with an unknown degree
of freedom. However, we can empirically determine it as a value
between 7 - 1 - 1 = 5 to 7 - 1 = 6." I can theoretically determine
the degrees of freedom here. There are 6 free parameters under the
alternative (since $q_0 + q_1 + q_2 + q_3 + q_4 + q_5 + q_6 = 1$), and
there are 3 free parameters under the null (since
$p_0 + p_1 + p_2 + p_3 = 1$), and so the degrees of freedom is 6 - 3 =
3, which is neither 5 nor 6. 

The degrees of freedom for the recursive test is incorrect in
@wang2022asymptotic. They say, right after their equation (1)
that the degrees of freedom is 6. But there are already 6 free
parameters under the alternative. Because @wang2022asymptotic
assume the double reduction rate is known, there is only 1 free
parameter under the null, the allele frequency. Thus, the true degrees
of freedom is 6 - 1 = 5, not 6. See Appendix S4 for an empirical
demonstration.

The degrees of freedom for the recursive test is incorrect in
@wang2021tracing. Right after their equation (3), they state
that their $\chi^2$ statistic "is thought to follow a chi-square
distribution with eight degrees of freedom." But there are already 8
parameters under the alternative (since $\sum_{k=0}^8q_k = 1$). The
number of parameters under the null is unclear since they are using a
different (incorrect) model for meiosis than I have studied for
octoploids, but it likely at least 1 (for the allele
frequency). Empirically, it seems the degrees of freedom is 7, not 8
(Appendix S4).


\clearpage

# Supplementary Figures

```{r alpha-box, fig.cap="Estimates of $\\alpha$ ($y$-axis) stratified by sample size ($x$-axis) and true $\\alpha$ (facets) using the method of Wang et al. (2022). The red dashed horizontal line is the true $\\alpha$ in each facet. The estimates are way off.", echo = FALSE, message = FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
paramdf <- read.csv("./sims.csv")
paramdf$n <- factor(paramdf$n)
paramdf$alphatext <- case_when(near(paramdf$alpha, 0) ~ "0",
                               near(paramdf$alpha, 1/7) ~ "1/7",
                               near(paramdf$alpha, 1/5) ~ "1/5",
                               near(paramdf$alpha, 3/11) ~ "3/11")
ggplot(paramdf, aes(x = n, y = alphahat)) +
  facet_wrap(. ~ alphatext) +
  geom_boxplot() +
  geom_hline(aes(yintercept = alpha), lty = 2, col = 2) +
  xlab("Sample Size") +
  ylab("Estimate of Double Reduction Rate")
```

```{r p-box, fig.cap="Estimates of $p_k$ ($y$-axis) for $k = 0,1,2,3$ (row facets) for different sample sizes ($x$-axis) and different initial values (truth or random) using the method of Wang et al. (2022). The red dashed horizontal line is the true $p_k$ in each facet. The estimates are somewhat biased.", echo = FALSE, fig.width=6}
gather(paramdf, key = "variable", value = "estimate", p0hat, p1hat, p2hat, p3hat) %>%
  mutate(variable = recode(variable, "p0hat" = "p0", "p1hat" = "p1", "p2hat" = "p2", "p3hat" = "p3")) ->
  cleandf

cleandf %>%
  select(variable, truth) %>%
  distinct() %>%
  mutate(value = case_when(truth == "A" ~ 0.25,
                           truth == "B" & variable == "p0" ~ 0.1,
                           truth == "B" & variable == "p1" ~ 0.2,
                           truth == "B" & variable == "p2" ~ 0.3,
                           truth == "B" & variable == "p3" ~ 0.4)) ->
  subdf

ggplot(cleandf, aes(x = n, y = estimate)) +
  facet_grid(truth ~ variable) +
  geom_boxplot() +
  geom_hline(data = subdf, mapping = aes(yintercept = value), lty = 2, col = 2)
```

\clearpage
    
# References
